% -------------------------------------------------------
\section{Conclusion}
% -------------------------------------------------------

This study examined how AI, ML, and NLP research discourse evolved over three decades using 24,677 arXiv abstracts and five complementary text analytics methods. Across all five methods, a coherent and mutually reinforcing account emerged.
The most consistent finding is the identification of 2014 - 2016 as a genuine inflection point, emerging independently from lexical analysis, topic modelling, classification, terminology diffusion tracking, and dimensionality reduction. The Pearson correlation of $r = 0.863$ between TF-IDF and Sentence-BERT centroid distance series two representations operating at entirely different levels of linguistic abstraction constitutes the strongest evidence that this shift is a real structural feature of the field rather than a methodological artefact. The shift is best characterized as expansion rather than replacement: theory-oriented topics declined but persisted, and most tracked terms were repositioned conceptually rather than disappearing. \newline
The analysis of how terminology spreads was the most original part of this study. By combining three methods, TF-IDF changes, Word2Vec changes in meaning, and SPECTER2 changes in discourse; the study was able to identify different patterns of adoption that no single method could show on its own. The most common pattern was subtle: terms changed in meaning or use without a big increase in how often they appeared. This happened for 13 out of the 36 terms studied. The example of “encoder" clearly shows this. Although its usage increased dramatically (by 63 times), its role in arguments completely shifted, something that only the SPECTER2 method could detect.
The pre-training countercase identified template replication as a structurally distinct pathway, a distinction that a frequency-only analysis would have missed entirely.\newline Limitation framing revealed a complementary dimension: ethics-related language grew from near-absence before 2015 to 13\% of abstracts by 2020–2024, with failure language becoming the second most prevalent framing category, shifts invisible to any undifferentiated count of limitation language.
Not all methods performed as intended. The limitation framing classifier measured broad subfield vocabulary shift rather than framing-specific change, and TF-IDF PCA recovered subfield composition variance rather than temporal drift. Both failures were diagnostically informative rather than simply negative, demonstrating the value of cross-method checking. \newline
Three challenges remain open. First, the SPECTER2 method does not capture some of the most important emerging terms like “transformer,” “prompt,” and “downstream task” because of how it is designed. Second, using only 6.9\% of the data as the Procrustes anchor may affect the results, meaning that some findings about changes in neighbourhoods could be influenced by the alignment method rather than real differences. Third, focusing only on abstracts may miss how limitations are discussed more fully in the main body of papers.
Overall, this suggests that no single method can fully explain how scientific discourse evolves. The strength of this study comes from comparing multiple approaches.

